Development of Physics or HPC Optimisation of a parallel 2D lattice Boltzmann solver using GPUs/CUDA

### 2017-08-09 Thesis/Draft.tex
more discussion

\subsection{2D model}
After initializing all the parameters, we need to initialize the new distribution functions and the new densities. This is done depending on the test case to be used, where each distribution function and density array are initialized according to the initial parameters and to the region of the fluid. The test cases are explained in more detail in Section \ref{sec:testcases}. \par
As mentioned in Section \ref{sec:cgModel}, the Color Gradient model imposes some changes on the sub-steps of the LBM solver. The most noticeable of theses changes happen in the collision sub-step. First, we need to calculate the color gradient of each node. Since this color gradient is the result of a sum of interactions from the 9 directions of the speed model, we need a separate loop before being able to use this array in the computation of the perturbation operator. However, the color gradient calculation depends on the orientation of the node, that is whether the node is in the north, south, east, west or middle of the mesh. We can then calculate the appropriate collision frequency and afterwards compute the perturbation, collision and recoloring operators by looping between the 9 directions. For the most part of this sub-step, parallelization is implemented by following the example of the solver's previous collision sub-step. Each node will have it's own thread to perform calculations, and since each node does not depend on it's neighbors we do not have to worry about conflicted access to memory. However, each thread needs to know it's orientation for the color gradient calculation. Instead of having each thread reverse calculate it's index in an x,y format, a new array was created and initialized in the initialization section of the solver. This array has an element per node and it's value represents the orientation of the node in the following manner:
\begin{itemize}
	\item 1 - North
	\item 2 - South 
	\item 3 - East 
	\item 4 - West 
	\item 0 - Middle
	\item -1 - Corner    
\end{itemize}
This way each thread simply needs to access this array with it's index to discover which orientation it needs to take into account. \par
The streaming step behaves like the original solver's streaming. We now have to stream not just one distribution function but both of them, as each fluid's changes are specific to it's own function. However, the 2D model was not behaving correctly when nodes near the boundary needed to be updated. This could be fixed by analyzing the mesh interpolation and how the stream array is built, which specifies whether a direction of the speed model should be updated or not. Since there was not enough time to perform this analysis, I used the same array as the one used in the color gradient to find out the orientation of node. Then I simply needed to specify how the streaming should work for each orientation.

\subsection{3D model}
\section{Verification and Validation}
\section{Test cases}\label{sec:testcases}
\section{Memory usage}
\section{Performance of the parallel code}\label{sec:performance}
\subsection{2D model}
\subsubsection{Normal distribution function v1.0}
\subsubsection{Enhanced distribution function v2.0}
\subsubsection{Optimizations v2.1}
\subsubsection{Loop unrolling v2.2}
\subsubsection{Final version v2.3}
\subsubsection{High order color gradient}
\subsubsection{CUDA compute capability}
\subsection{3D model}
\subsubsection{Enhanced distribution function v1.0}
\subsubsection{Final version v1.1}
\subsubsection{High order color gradient}
\subsubsection{CUDA compute capability}
\subsection{Hardware}
\subsection{Double vs Single}
\chapter{Conclusions}

>>>>

\subsection{2D model}
After initializing all the parameters, we need to initialize the new distribution functions and the new densities. This is done depending on the test case to be used, where each distribution function and density array are initialized according to the initial parameters and to the region of the fluid. The test cases are explained in more detail in Section \ref{sec:testcases}. \par
As mentioned in Section \ref{sec:cgModel}, the Color Gradient model imposes some changes on the sub-steps of the LBM solver. The most noticeable of theses changes happen in the collision sub-step. First, we need to calculate the color gradient of each node. Since this color gradient is the result of a sum of interactions from the 9 directions of the speed model, we need a separate loop before being able to use this array in the computation of the perturbation operator. However, the color gradient calculation depends on the orientation of the node, that is whether the node is in the north, south, east, west or middle of the mesh. We can then calculate the appropriate collision frequency and afterwards compute the perturbation, collision and recoloring operators by looping between the 9 directions. For the most part of this sub-step, parallelization is implemented by following the example of the solver's previous collision sub-step. Each node will have it's own thread to perform calculations, and since each node does not depend on it's neighbors we do not have to worry about conflicted access to memory. However, each thread needs to know it's orientation for the color gradient calculation. Instead of having each thread reverse calculate it's index in an x,y format, a new array was created and initialized in the initialization section of the solver. This array has an element per node and it's value represents the orientation of the node in the following manner:
	\item 1 - North
	\item 2 - South 
	\item 3 - East 
	\item 4 - West 
	\item 0 - Middle
	\item -1 - Corner    
This way each thread simply needs to access this array with it's index to discover which orientation it needs to take into account. \par
The streaming step behaves like the original solver's streaming. We now have to stream not just one distribution function but both of them, as each fluid's changes are specific to it's own function. However, the 2D model was not behaving correctly when nodes near the boundary needed to be updated. This could be fixed by analyzing the mesh interpolation and how the stream array is built, which specifies whether a direction of the speed model should be updated or not. Since there was not enough time to perform this analysis, I used the same array as the one used in the color gradient to find out the orientation of node. Then I simply needed to specify how the streaming should work for each orientation. \par
The boundary step also behaves like the original solver's boundary update. However, the tests used to validate the Color Gradient model mostly use periodic boundaries, which did not exist in the 2D version of the previous solver. As such, these boundaries needed to be implemented and added to the 2D version of the solver. Furthermore, the boundaries also need an interpolation of the mesh in order for each thread to know what type of boundary that node should be updated with. Since the lattice generator software was out of scope for this thesis, I again used the same array built with the orientations for the color gradient model. This way, each thread could identify which boundary condition should be applied depending on the position of the node. \par
The macro variable updates need to be replicated for both distribution functions and densities. Also, the calculation of the velocity needs to take both fluids into account. Other than that, it behaves in the same manner as the original solver. \par
Finally, the residuals calculation also needed to be updated since the previously implemented residuals based on the sum of absolute differences between each iteration's distribution function was not properly obeying the convergence criteria for our multiphase approach. Now, the residuals are based on the maximum difference of the distribution function and the user can choose which should the convergence criteria be in the setup file.

\subsection{3D model}
The 3D model adaptation was much easier than the 2D implementation since for the most part it was just a matter of extending the computations on the distribution functions, from 9 to 19 directions. The core logic behind every operation was the same. However, the collision model needed to be updated since the collision frequencies, perturbation operator and equilibrium distribution function calculation were based on a enhanced version of the 2D model's version. Eventually we also modified the 2D version to use this enhanced collision step since it was meant to provide more accurate results. \par
One of the most significant improvements from implementing the 3D model was that the previous solver had more enhancements in 3D than in 2D. We could now use the proper streaming and boundary steps since there were no complications when generating 3D meshes and the software could adequately interpolate these meshes and initialize control arrays correctly.

\section{Validation}
The validation of the results was performed by Antonio González \cite{antonioThesis}. To do so, every test case was implemented according to their specific needs. Almost every case was based on the same mesh, a simple square or cube depending on the dimension. The only exception was the Rayleigh-Taylor instability where a rectangular mesh was used. Appendix \textcolor{red}{Ref appendix validation} shows that the model produces correct results for every test case. \par
The solver has the option of producing results in different formats, .dat, .vti and .csv. .VTI files were the addition of last year's authors and can be read using Paraview \cite{paraview}. The main advantage of these files is that they require significantly less memory than the other formats. Because of this, we are able to save time by transferring the results from Cranfield's HPC center faster and also by making post-processing analysis less time consuming. For this reason, we extended the creation of these files to use the new multiphase model arrays.
\section{Test cases}\label{sec:testcases}
The validation of the CG model depends not only on the analysis of the actual result of the simulation, but also in type of test case used. Each test case had different requirements in the software since they depended on different arrays and each used their own type of verification. Because of this, I added the option to change between test 
\section{Memory usage}
\section{Performance of the parallel code}\label{sec:performance}
\subsection{2D model}
\subsubsection{Normal distribution function v1.0}
\subsubsection{Enhanced distribution function v2.0}
\subsubsection{Optimizations v2.1}
\subsubsection{Loop unrolling v2.2}
\subsubsection{Final version v2.3}
\subsubsection{High order color gradient}
\subsubsection{CUDA compute capability}
\subsection{3D model}
\subsubsection{Enhanced distribution function v1.0}
\subsubsection{Final version v1.1}
\subsubsection{High order color gradient}
\subsubsection{CUDA compute capability}
\subsection{Hardware}
\subsection{Double vs Single}
\chapter{Conclusions}
